---
title: "Binomial Distribution in Epidemiology"
id: "binomial-demo-se01-sec01-sp01"
section_id: "binomial-demo-se01-sec01"
session_id: "binomial-demo-se01"
format:
  html:
    toc: true
    toc-depth: 2
execute:
  echo: true
  warning: false
  message: false
---

## Overview

The binomial distribution is a simple but powerful way of modelling situations where we observe a fixed number of independent trials, each with two possible outcomes. In epidemiology, it can be useful when we want to estimate the probability of observing a certain number of cases in a small sample under a known or assumed risk.

::: {.callout-important}
## Why this matters

The binomial distribution provides a practical bridge between probability theory and real epidemiological reasoning. It helps learners connect mathematical concepts to questions such as: *How likely is it that exactly 3 out of 10 individuals test positive if the risk of infection is 0.2?*
:::

## Learning goals

By the end of this page, you should be able to:

- explain the key assumptions of the binomial distribution
- interpret a simple epidemiological example
- read R code used to calculate binomial probabilities
- interpret a visualisation of the resulting probability distribution

## The core idea

The binomial distribution is appropriate when all of the following apply:

1. there is a fixed number of trials
2. each trial has two possible outcomes
3. the probability of the event stays constant
4. trials are independent

In an epidemiology context, a “trial” might be one person in a sample, and the two outcomes might be “infected” or “not infected”.

## Contexts for understanding the model

::: {.panel-tabset}

## Mathematical definition

A binomial random variable gives the probability of observing exactly \(x\) events in \(n\) trials when the probability of the event is \(p\).

\[
P(X=x)=\binom{n}{x}p^x(1-p)^{n-x}
\]

This tells us the probability of getting exactly a specific number of successes.

## Epidemiology interpretation

In epidemiology, the “success” is not necessarily something positive. It simply means the event of interest happened.

For example:

- \(n = 10\) individuals
- \(p = 0.2\) probability of infection for each individual
- \(x = 3\) infected individuals observed

The model estimates the probability of seeing exactly 3 infections in that sample.

## Practical example

Suppose a small screening exercise is carried out with 10 individuals, and each has an independent probability of infection of 0.2.

A binomial model can help answer questions such as:

- What is the probability of seeing exactly 3 infections?
- What is the most likely number of infections?
- How unusual would 6 infections be?
:::

## Watch and notice

The short video below introduces the binomial distribution. As you watch, focus on these questions:

- What are the required assumptions?
- What do “number of trials” and “probability of success” mean?
- How might this relate to infection or case detection in epidemiology?

{{< video https://www.youtube.com/watch?v=ATUcJmGxJqs >}}

## R code example

Below is a simple R example using an epidemiological scenario.

We assume:

- 10 independent individuals
- probability of infection = 0.2
- interest in the probability of exactly 3 infections

```{r}
n <- 10
p <- 0.2

# Probability of exactly 3 infections
prob_exact_3 <- dbinom(3, size = n, prob = p)
prob_exact_3
```

The output gives the probability of observing exactly 3 infections in this sample under the assumptions of the model.

## Full distribution across the sample

We can also calculate the full distribution of possible outcomes, from 0 to 10 infections.

```{r}
x <- 0:n
y <- dbinom(x, size = n, prob = p)

binom_df <- data.frame(
  infections = x,
  probability = y
)

binom_df
```

## Visualising the distribution

The plot below shows the probability of observing each possible number of infections.

```{r}
library(plotly)

plot_ly(
  data = binom_df,
  x = ~infections,
  y = ~probability,
  type = "bar"
) |>
  layout(
    title = "Binomial distribution: infections in a sample of 10",
    xaxis = list(title = "Number of infections"),
    yaxis = list(title = "Probability")
  )
```

## Interpreting the visualisation

The tallest bar shows the most likely number of infections under this model. Values further away from the centre become less likely.

This kind of plot helps students see that the model is not just about one answer. It describes a full range of possible outcomes and their probabilities.

## Reveal: when the model may not fit well

::: {.callout-note}

## Before you open this

Try to answer this first:

**Why might a binomial model not always be appropriate in real epidemiology?**
:::

::: {.callout-tip collapse="true"}

## Reveal answer

The binomial model depends on assumptions that may not hold in real-world epidemiology.

For example:

* infections may not be independent if people are connected through households, workplaces, or schools
* the probability of infection may not be constant across individuals
* exposure risk may vary over time or by subgroup
* outbreaks often involve clustering rather than random independence

This means the binomial distribution is useful for teaching and for some simple problems, but it is not always a realistic model of transmission dynamics.
:::

## Quick self-check

Consider the following question:

If the probability of infection stays at 0.2 and the sample size stays at 10, would observing 8 infections seem likely under this model?

Write a one-sentence answer before checking the plot above.

## Interpretation prompt

What does the shape of the distribution suggest about the most plausible number of infections in this scenario? How would your interpretation change if the probability increased from 0.2 to 0.5?

## Summary

This page demonstrates how the binomial distribution can be taught as part of a realistic epidemiology learning experience.

It combines:

* concept explanation
* contextual interpretation
* video-based learning
* executable R code
* interactive data visualisation
* reflective and interpretive prompts

In doing so, it shows how a structured publishing workflow can generate a rich, student-facing learning environment rather than a static content page.

<!-- START_NAVIGATION -->

```{=html}
<div class="custom-page-navigation">
    <div class="nav-prev">
        <a href="index.qmd">
            <span class="nav-label">← Previous</span>
            <span class="nav-title">Understanding the Binomial Distribution</span>
        </a>
    </div>
    <div class="nav-next">
    </div>
</div>
```

<!-- END_NAVIGATION -->
